## Assessment of hypothesis 3

### Methods:

#### Section 2.2.: Assessment of the proposed hypotheses

To assess $H_3$ we first identified all DEM cells reachable within five hours from the Bau, temporarily
adding each as a node (hereafter referred to as node x) to the network of sources. We chose a five-hour
cut-off somewhat arbitrarily, to keep computations manageable, but preliminary findings suggest that
adding more distant locations would not have been informative (see section 3.4.). For each cell we then
evaluated all valid procurement paths to the Bau (i.e., $v_{x→v→b}$ for each cell x, and for each v'
representing an accessible raw material source; see Fig. 1D), identifying optimal alternatives (i.e.,
highest $P_{S1 (v’)}$ value per cell) as well as ‘good’ alternatives whose $P_{S1 (v’)}$ values fall within the 95%
confidence interval of the highest, to account for estimation uncertainties. We examined the
proportions of $S_0$ sources among the ‘good’ alternatives at each DEM cell across the region, to
determine the degree to which different landscape locations are compatible with $H_3$, and investigated
any spatial patterning. Finally, for each raw material type we also summed the number of DEM cells
where their sources are identified as good or optimal procurement targets, that is, we calculated the
areal extent over which a hominin could be expected to have targeted a given material, and tested for a
correlation with the frequency (i.e., number of lithics) with which the different raw materials are
represented at the site.


```{r load-mod, echo=FALSE, message=FALSE, include=FALSE, results="asis"}
knitr::opts_chunk$set(echo = FALSE)
library(rgdal)
library(raster)
library(rasterVis)

source("../R/model_main.R")
```

```{r echo = FALSE, include = FALSE}
# Load data:

#source("../R/gis_h3_datagen.R")
#source("../R/gis_h3_datagen_aux.R") # Generates eval_hodes_5h.rds
eval_nodes <- readRDS("../data/eval_nodes_5h.rds")

# Load data required for figure (next block)
 # Blank Bau costmap
bau_reg <- readRDS("../data/bau_classif_blank_fullext_verified.rds")
wt_contours <- readOGR("../data/time_fromBau_contours", verbose = FALSE)
sites_v <- readRDS("../data/bau_coor.rds")
sources_v <- readOGR("../data/sources_utm_abr", verbose = FALSE)
aid_chulls <- readRDS("../data/aids.chulls.rds")
cell_size <- bau_reg@grid@cellsize[1] * bau_reg@grid@cellsize[1] # m2

# Compute PS1 values with confidence intervals:
p_fits <- predict.glm(object = mod_ns, newdata = eval_nodes,
                      type = "link", se.fit = TRUE)

eval_nodes$targ_ps1 <- exp(p_fits$fit) / (1 + exp(p_fits$fit))
eval_nodes$targ_ps1_lwr <- p_fits$fit -
                           p_fits$se.fit * abs(qt(p = 0.025,
                                                  df = mod_ns$df.residual))
eval_nodes$targ_ps1_lwr <- exp(eval_nodes$targ_ps1_lwr) /
                               (1 + exp(eval_nodes$targ_ps1_lwr))
eval_nodes$targ_ps1_upr <- p_fits$fit +
                           p_fits$se.fit * abs(qt(p = 0.025,
                                                  df = mod_ns$df.residual))
eval_nodes$targ_ps1_upr <- exp(eval_nodes$targ_ps1_upr) /
                               (1 + exp(eval_nodes$targ_ps1_upr))

# Reduce data to alternatives that fall within the 95CI for each cell:
eval_nodes[, max_targ_ps1_lwr := max(targ_ps1_lwr), by = "cell"]
eval_nodes_red <- eval_nodes[targ_ps1 > max_targ_ps1_lwr]

eval_nodes_red <- merge(eval_nodes_red, glm_sel[, c("sid", "aid")],
                        by.x = "target_sid", by.y = "sid")
eval_nodes_red$t.pot.used <- eval_nodes_red$target_sid %in% glm_sel[pot.used == 1]$sid

# Calculate per node stats:
eval_nodes_sum <- eval_nodes_red[, list(nr_targs = .N,
                              nr_rep = length(which(t.pot.used == 1)),
                              nr_nonrep = .N - length(which(t.pot.used == 1))),
                            by = "cell"]
eval_nodes_sum$perc_rep <- round(eval_nodes_sum$nr_rep /
                                 eval_nodes_sum$nr_targs * 100)

eval_nodes_sum$usage_code <- factor(levels = c("100%", "75% to 99%",
                                               "50% to 74%", "25% to 49%",
                                               "1% to 24%", "0%"),
                              ordered = TRUE)
eval_nodes_sum[perc_rep == 100]$usage_code <- "100%"
eval_nodes_sum[perc_rep >= 75 & perc_rep < 100]$usage_code <- "75% to 99%"
eval_nodes_sum[perc_rep >= 50 & perc_rep < 75]$usage_code <- "50% to 74%"
eval_nodes_sum[perc_rep >= 25 & perc_rep < 50]$usage_code <- "25% to 49%"
eval_nodes_sum[perc_rep > 0 & perc_rep < 25]$usage_code <- "1% to 24%"
eval_nodes_sum[perc_rep == 0]$usage_code <- "0%"

# Aggregated statistics on targetted sources and SAs:
sid_hits <- eval_nodes_red[, list(freq = .N), by = "target_sid"]
aid_hits <- eval_nodes_red[, list(freq = .N), by = "aid"]
aid_hits <- merge(aid_hits,
                 glm_sel[, list(maxL = max(lithics, na.rm = TRUE)), by = "aid"],
                 by.x = "aid", by.y = "aid")
aid_hits_s1 <- aid_hits[maxL > 0]

# Compute areas serviced by different 'sources'good' alternatives:
eval_nodes_areas <- eval_nodes_red[, .N, by = "target_sid"]
eval_nodes_areas$t.pot.used <- eval_nodes_areas$target_sid %in% glm_sel[pot.used == TRUE]$sid
eval_nodes_areas$km2 <- eval_nodes_areas$N * cell_size / (1000 * 1000)

cor_good_all <- cor.test(aid_hits$maxL, aid_hits$freq, method = "spearman",
                        exact = FALSE)
cor_good_s1 <- cor.test(aid_hits_s1$maxL, aid_hits_s1$freq, method = "spearman",
                       exact = FALSE)

# Optimal alternatives per cell, and correlations with lithics
eval_nodes_opt <- eval_nodes_red[, .SD[which.max(targ_ps1)], by = "cell"]
eval_nodes_opt_red <- merge(glm_sel[!is.na(lithics), c("aid", "lithics")],
                            eval_nodes_opt[, .N, by = "aid"], by = "aid")
cor_opt_red <- cor.test(eval_nodes_opt_red[aid != 58]$lithics,
                        eval_nodes_opt_red[aid != 58]$N, method = "spearman")
cor_opt_all <- cor.test(eval_nodes_opt_red$lithics, eval_nodes_opt_red$N,
                        method = "spearman")

```

### Results

#### 3.5. Evaluation of Hypothesis 3

<!-- Figure 9 code (plot at the end of the file) -->
```{r res-h3, include = FALSE}

# Populate raster with classified data:
bp_dt <- data.table(val = bau_reg@data$coston_SimRegion_inv,
                    cell = rownames(bau_reg@data))
bp_dt$cell <- as.numeric(bp_dt$cell)
setkeyv(bp_dt, "cell")
setkeyv(eval_nodes_sum, "cell")
bp_comb <- merge(bp_dt, eval_nodes_sum, all = T)
bau_reg@data$coston_SimRegion_inv <- bp_comb$usage_code
#writeRAST(bau_reg, "../data/classif_usage_byloc_rawmodel.tif")


# Crop data
bau_reg_rast <- raster::raster(bau_reg)
bau_reg_rast <- crop(bau_reg_rast, extent(wt_contours))
sources_v_c <- sources_v[which(!is.na(extract(bau_reg_rast, sources_v))), ]

# Transform to a data.table obj and add extra info:
sources_v_c <- data.table(sources_v_c@data[, c("SourceID", "Elevation",
                                              "x", "y", "z")])
names(sources_v_c)[1] <- "sid"

gis_sel_c <- glm_sel # Work on a copy
gis_sel_c[, maxL := max(lithics, na.rm = TRUE), by = "aid"] # Max proc. from src

# Remember that targets MAY be outside of zone!
sources_v_c <- merge(sources_v_c, sid_hits,
                     by.x = "sid", by.y = "target_sid", all = TRUE)
sources_v_c <- merge(sources_v_c,
                     gis_sel_c[, c("sid", "pot.used", "maxL")],
                     by.x = "sid", by.y = "sid")
sources_v_c[is.na(sources_v_c$freq)]$freq <- 0

sources_v_c_s1 <- sources_v_c[pot.used == 1]
sources_v_c_s0 <- sources_v_c[pot.used == 0]
coordinates(sources_v_c_s1) <- 3:4
coordinates(sources_v_c_s0) <- 3:4

sites_v_c <- sites_v # Originally this read a Shapefile. Keep for now.
coordinates(sites_v_c) <- 6:7

# **Exploited** source area polygons:
p <- lapply(aid_chulls, sp::Polygon)
aid_chulls_poly <- sp::SpatialPolygons(lapply(seq_along(p),
                          function(i) Polygons(list(p[[i]]), ID = names(p)[i])))

# PLOT:
cols <- c("#4DAF4A", "#8DD3C7", "#80B1D3", "#BEBADA", "#FDB462", "#FB8072")
myTheme <- rasterTheme(region = cols)

g <- levelplot(bau_reg_rast, par.settings = myTheme) +
  latticeExtra::layer(sp.polygons(wt_contours, lwd = 2)) +
  latticeExtra::layer(sp.points(sources_v_c_s1, pch = 19,
                                col = rgb(0, 1, 0, 0.6),
                          cex = 0.7 + log(sources_v_c_s1@data$freq + 1) / 7)) +
  latticeExtra::layer(sp.points(sources_v_c_s1, pch = 21, col = "black",
                          cex = 0.7 + log(sources_v_c_s1@data$freq + 1) / 7)) +
  latticeExtra::layer(sp.points(sources_v_c_s0, pch = 18,
    col = rgb(1, 0, 0, 0.6), cex = 1 + log(sources_v_c_s0@data$freq + 1) / 7)) +
  latticeExtra::layer(sp.points(sources_v_c_s0, pch = 9, col = "black",
                           cex = 0.6 + log(sources_v_c_s0@data$freq + 1) / 7)) +
  latticeExtra::layer(sp.points(sites_v_c[which(sites_v_c$Name == "Bau")],
                                pch = 13, cex = 5, lwd = 2, col = "black")) +
  latticeExtra::layer(sp.polygons(aid_chulls_poly, line = "black"))

png("../plots/Fig_9.png", height = 10, width = 15, units = "in", res = 600)
g
dev.off()

```

The set of sources predicted to be ignored or avoided under Hypothesis 3 ($H_3$), and
consequently the set of sources deemed to have been exploitable, is identical to that predicted under $H_2$
for any given maximum cut-off value for travel times from the Bau (see section 3.4.). When this cut-off is
set to five hours, `r length(unique(eval_nodes_opt[t.pot.used == TRUE]$target_sid))` $S_1$
sources are identified as optimal alternatives over `r round(nrow(eval_nodes_opt[t.pot.used == TRUE]) / nrow(eval_nodes_opt) * 100, 1)`% of the area (ca. `r round(nrow(eval_nodes_opt[t.pot.used == TRUE]) * cell_size / (1000 * 1000))` $km^2$
of `r round(nrow(eval_nodes_opt) * cell_size / (1000 * 1000))` $km^2$),
and a lone (`r length(unique(eval_nodes_opt[t.pot.used == FALSE]$target_sid))`) $S_0$ source (ID
`r unique(eval_nodes_opt[t.pot.used == FALSE]$target_sid)`) is identified as the optimal procurement choice over the
remaining `r round(nrow(eval_nodes_opt[t.pot.used == FALSE]) / nrow(eval_nodes_opt) * 100, 1)`% -
in other words, $H_3$, like $H_2$, can account for the lack of utilization of
`r round((nrow(glm_sel[time_fromBau <= 5 * 3600 & pot.used == 0]) - 1) / nrow(glm_sel[time_fromBau <= 5 * 3600 & pot.used == 0]) * 100, 1)`% of the `r nrow(glm_sel[time_fromBau <= 5 * 3600 & pot.used == 0])`
unused ($S_0$) sources available within five hours of the site. The number of locations (i.e., DEM cells)
where these $S_1$ and $S_0$ optimal alternatives are expected to have been targeted has a predictable but not
statistically significant relationship with the number of lithics made from materials that can be procured
at these, with one exception (Spearman’s $r_s$ [`r nrow(eval_nodes_opt_red[aid != 58]) - 2`] = `r round(as.numeric(cor_opt_red$estimate), 2)`, p = `r round(as.numeric(cor_opt_red$p.value), 3)` with source ID `r unique(eval_nodes_opt[aid == 58]$target_sid)` excluded; $r_s$ [`r nrow(eval_nodes_opt_red) - 2`] = `r round(as.numeric(cor_opt_all$estimate), 2)`,
p = `r round(as.numeric(cor_opt_all$p.value), 3)`
with it included). This exception, which is identified as an optimal choice over ca. `r round(eval_nodes_opt_red[aid == 58]$N / nrow(eval_nodes_opt) * 100, 1)`% of the
area, only accounts for up to `r glm_sel[lithics > 0 & aid == 58]$lithics` of the
provenanced lithics found at the site, and may be explainable by
a lack of utilization of certain areas or, possibly, by an incorrect (over-)estimation of its $P_{S1(v')}$ value. Be
that as it may, optimal resource selection based on knowledge of the relative benefits afforded by the
available sources (i.e., nodule sizes, quality of the materials, and extent over which they may be found)
and the cost of accessing them, fit the data well, though not necessarily equally well, regardless of
where a hominin might have been located when deciding where to procure raw materials on their way
to the Bau.



The set of ‘good’ alternatives identified for locations (i.e., DEM cells) within five-hours of the Bau
includes `r length(unique(eval_nodes_red$target_sid))` unique sources. Of these,
`r length(unique(eval_nodes_red[t.pot.used == TRUE]$target_sid))` belong
to the set $S_1$ that yield archaeologically exploited raw
materials, representing `r length(unique(eval_nodes_red[t.pot.used == TRUE]$aid))` of
the utilized source areas and accounting for up to `r sum(glm_sel[aid %in% unique(eval_nodes_red[t.pot.used == TRUE]$aid)]$lithics, na.rm = TRUE)` of the
provenanced lithics (i.e., `r round(sum(glm_sel[aid %in% unique(eval_nodes_red[t.pot.used == TRUE]$aid)]$lithics, na.rm = TRUE) / sum(glm_sel$lithics, na.rm = TRUE) * 100, 1)`%) at the site. The other `r length(unique(eval_nodes_red[t.pot.used == FALSE]$target_sid))` are sources that are not represented at the
Bau, and set an upper limit to the number of seemingly unused sources (of the `r nrow(glm_sel[time_fromBau <= 5 * 3600 & pot.used == 0])` available within five
hours) that cannot be explained under $H_3$. The number of ‘good’ alternatives identified per DEM cell
varies (mean = `r round(mean(eval_nodes_sum$nr_targs), 1)`, SD = `r round(sd(eval_nodes_sum$nr_targs), 1)`, maximum = `r max(eval_nodes_sum$nr_targs)`) and, as shown in Figure 9, they typically consist of $S_1$ sources (mean =
`r round(mean(eval_nodes_sum$perc_rep), 1)`%, minimum = `r round(min(eval_nodes_sum$perc_rep), 1)`%, maximum = `r round(max(eval_nodes_sum$perc_rep), 1)`%). At most locations (`r round(nrow(eval_nodes_sum[perc_rep > 50])/ nrow(eval_nodes_sum) * 100, 1)`%), these constitute half or
more of the alternatives, and no ‘good’ $S_0$ options can be identified over an area of some `r round(nrow(eval_nodes_sum[perc_rep == 100]) * cell_size / (1000 * 1000))` $km^2$ (`r round(nrow(eval_nodes_sum[perc_rep == 100])/ nrow(eval_nodes_sum) * 100, 1)`%).

Digital elevation model cells where specific sources are identified as ‘good’ options form areas of
variable size (mean = `r round(mean(eval_nodes_areas$km2), 1)` $km^2$, 
median = `r round(median(eval_nodes_areas$km2), 1)` $km^2$, maximum = `r round(max(eval_nodes_areas$km2), 1)` $km^2$).
A Spearman’s rank
correlation test indicates that when these areas are summed for sources of each of the
`r length(unique(eval_nodes_red$aid))` raw material types
represented by the `r length(unique(eval_nodes_red$target_sid))` unique options,
they are moderately and positively correlated with the number of
sourced artifacts made from those materials ($r_s$ [`r nrow(aid_hits) - 2`] = `r round(as.numeric(cor_good_all$estimate), 2)`, p = `r round(as.numeric(cor_good_all$p.value), 3)`). If we assume the identification of
non-utilized sources among the ‘good’ alternatives represents an error, and therefore only consider the
areas serviced by alternatives from exploited source areas, the correlation becomes much stronger ($r_s$ 
[`r nrow(aid_hits_s1) - 2`] = `r round(as.numeric(cor_good_s1$estimate), 3)`, p = `r round(as.numeric(cor_good_s1$p.value), 3)`).
We interpret these correlations as indicating that 1) in the aggregate, and based
solely on the lithic data, we cannot reject the possibility of a relatively uniform utilization of the region
by the hominins responsible for the accumulation of the Bau assemblages; and 2) that, regardless of
where they might have been located, said hominins could identify and target sources that would have
been optimal or close to optimal procurement choices.

```{r fig.cap="Figure 9. Predicted resource selection across the landscape. The different colors indicate the proportions of good procurement options consisting of raw material (RM) sources from exploited source areas (i.e., S1 sources), as computed for every location (i.e., digital elevation model [DEM] cell; 80 m resolution) under Hypothesis 3. Locations (i.e., DEM cells) where these proportions reach minimal (1–24%) and maximal (100%) values are shown in orange and dark green respectively, while cells with intermediary values are shown in purple, blue and teal. Green circles represent S1 sources available within five hours of the site, while red diamonds represent S0 (i.e., unused) sources; the size of the circles and diamonds indicates the size of the area (i.e., number of DEM cells) over which the respective sources are identified as good procurement alternatives, on a logarithmic scale. Green polygons represent convex hulls encompassing sources from individual exploited source areas, and concentric rings show GIS-computed distances that can be covered walking away from the site, in one-hour increments (minimum walking times). Note that throughout the region the identified good procurement options are mostly sources from exploited source areas; some unused sources (red diamonds) located close to the site are identified as good procurement alternatives from a large number of locations, but from those same locations good alternatives also include S1 sources, the latter accounting for a majority of options in most cases."}

g

```
