## Logistic resource selection model:

```{r load-mod, echo=FALSE, message=FALSE, include=FALSE, results="asis"}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(car)
source("../R/model_main.R")
source("../R/cmp_stat_fcns.r")
```

### Results:

#### 3.1. Resource selection model and PS1 estimation

A null model comparison indicates that the evaluated characteristics of the available sources have a
highly significant combined influence on their probabilities ($P_{S1}$) of yielding raw material types that are
also archaeologically represented at the Bau ($\chi^2$=`r round(res_mod_sig$Deviance[2], 2)`,
df=`r res_mod_sig$Df[2]`, p `r report_pvals(res_mod_sig$P[2])`). As shown in Table 2<!-- \@ref(tab:som-moddiag-infl-coefs) --> (see also
SOM S4), the quality of the raw materials, the size of the area over which these are found, and the
abundance of large nodules, all have a positive effect. However, the strongest and most significant
predictor is the minimum walking time required to reach the sources from the Bau, and its effect is
negative; the higher the access costs of a source, the less likely it is to provide raw material types
represented at the site. These results are consistent with previous findings (e.g., Browne and Wilson,
2011) as well as theoretical expectations.

```{r res-mod-coefs, message=FALSE}

p_vals <- unlist(lapply(coef(summary(mod))[, "Pr(>|z|)"], report_pvals))
round_to <- 3
mod.sum <- data.frame("Term (scaled)" = c("Intercept", "Quality (log)",
                                        "Extent (sqrt)", "Time from Bau (sqrt)",
                                        "Large rock abundances"),
                "Estimate" = round(coef(summary(mod))[, "Estimate"], round_to),
                "SE" = round(coef(summary(mod))[, "Std. Error"], round_to),
                "Lower CL" = round(confint(object = mod)[, 1], round_to),
                "Upper CL" = round(confint(object = mod)[, 2], round_to),
                "Unscaled mean" = c("N/A",
                                    round(mean(glm_sel$qual.log), round_to),
                                    round(mean(glm_sel$ext.sqrt), round_to),
                                    round(mean(glm_sel$time.sqrt), round_to),
                                    round(mean(glm_sel$large_r), round_to)),
                      "Unscaled SD" = c("N/A",
                                        round(sd(glm_sel$qual.log), round_to),
                                        round(sd(glm_sel$ext.sqrt), round_to),
                                        round(sd(glm_sel$time.sqrt), round_to),
                                        round(sd(glm_sel$large_r), round_to)),
                      "P" = p_vals)

knitr::kable(mod.sum,
                       col.names = c("Term (scaled)", "Estimate", "SE",
                                   "Lower CL", "Upper CL", "Mean", "SD", "P"),
                       row.names = FALSE,
                       format = "html",
                       caption = "Table 2. Scaled logistic resource selection
            model coefficients with unscaled means and standard deviations. ")

```



The estimated $P_{S1}$ values for individual raw material sources are quite variable (min = `r round(min(glm_sel$probs.full), 2)`,
max = `r round(max(glm_sel$probs.full), 2)`, mean = `r round(mean(glm_sel$probs.full), 2)`, SD = `r round(sd(glm_sel$probs.full), 2)`), and this is true even of sources assigned to the same source
area. Estimated probabilities for the seven locations assigned to source area 55, for example, which
yields the most common raw material type in the Bau assemblage (`r max(glm_sel[aid == 55]$lithics, na.rm=T)` artifacts), vary between a
minimum of `r round(min(glm_sel[aid == 55]$probs.full), 2)` and a maximum of `r round(max(glm_sel[aid == 55]$probs.full), 2)`, having a mean of `r round(mean(glm_sel[aid == 55]$probs.full), 2)`. This variability is expected given the
substantial differences in the characteristics of the sources (see SOM S2), including access costs (see Fig.
2) and the overall quality of the available nodules. It is nevertheless noteworthy because it strongly
suggests that not all locations within a given source area represented at the Bau were exploited to the
same extent and, indeed, some may not have been utilized at all. Also noteworthy, particularly given our
present goals, is that several sources which are not represented at the Bau (n = `r nrow(glm_sel[pot.used == 0 & probs.full > .5])`, or `r round(nrow(glm_sel[pot.used == 0 & probs.full > .5])/nrow(glm_sel[pot.used == 0]) * 100, 0)`%) were assigned
high probabilities by our model ($P_{S1}$ >= `r min(glm_sel[pot.used == 0 & probs.full > .5]$probs.full)`; max = `r max(glm_sel[pot.used == 0 & probs.full > .5]$probs.full)`).

### SOM S3 - Logistic model diagnostics

An examination of variance inflation factors (VIFs) revealed that collinearity
was not an issue (maximum VIF = `r round(max(car::vif(mod)), 2)`).
Despite the imbalance in the frequency of the two response alternatives, the model
also appears to be stable, as case-wise deletions do not substantially affect
individual estimated coefficients (SOM Table S6; see also SOM Fig. S4 for standardized dfbeta values,
which indicate the standardized difference between coefficient estimates with cases excluded one at a time).


``` {r som-moddiag}
mod_dfb <- data.frame(slope = mod$coefficients,
                      mod$coefficients + t(apply(X = dfbeta(mod),
                                                 MAR = 2, FUN = range)))
mod_dfb <- round(mod_dfb, 4)
mod_dfb <- cbind(rownames(mod_dfb), mod_dfb)
names(mod_dfb) <- c("Coefficient", "Slope estimate", "Minimum", "Maximum")
rownames(mod_dfb) <- 1:nrow(mod_dfb)
knitr::kable(mod_dfb, caption = "SOM Table S6. Baseline scaled (z-transformed)
             logistic model coefficients with minimum and maximum ranges with
             case-wise deletions.")
```

```{r som-moddiag-plot, fig.cap= "SOM Figure S4. Plot of standardized dfbeta values for individual scaled predictors"}
car::dfbetasPlots(mod)
jpeg(filename = "../plots/dfbetas.jpg", width = 3000, height = 2000,
     unit = "px", quality = "1200", pointsize = 70)
car::dfbetasPlots(mod, lwd = 2)
dev.off()

```

```{r}
lev_thresh <- 2 * (length(coefficients(mod)) + 1) / length(residuals(mod))
```

We did identify `r length(which(hatvalues(mod) > lev_thresh))` influential cases (SOM Table S7), defined
here as having leverage values above a threshold of two times the
number of predictors plus one divided by the number of cases (i.e.
2*`r length(coefficients(mod)) + 1`/`r length(residuals(mod))`, or
`r round(lev_thresh, 3)`). However, removal of these influential cases does not
have a major effect on the results: all coefficients remain within
their standard error as observed in the full model with no cases removed, no coefficients
change signs and no critical changes in the significance of individual predictors
can be observed (see SOM Table S8). Most affected by the
removal of influential cases is the estimate for the quality variable, which is
not surprising given that the removed cases have some of the highest quality
values in the dataset.

```{r som-moddiag-infl-tab}

data_infl <- cbind(glm_sel[which(hatvalues(mod) > lev_thresh),
                          c("sid", "aid", "pot.used", "time_fromBau", "quality",
                            "extent", "large_r")],
              data.frame(leverage = round(hatvalues(mod)[which(hatvalues(mod) >
                                                              lev_thresh)], 4)))
data_infl$time_fromBau <- round(data_infl$time_fromBau / 3600, 1)
data_infl$extent <- round(data_infl$extent)
data_infl$large_r <- as.character(data_infl$large_r)

pot.used.srcs <- which(data_infl$pot.used == 1)
data_infl$pot.used <- as.character(data_infl$pot.used)
data_infl[pot.used.srcs, ]$pot.used <- "Potentially utilized"
data_infl[-pot.used.srcs, ]$pot.used <- "Not used"
data_infl[which(data_infl$large_r == "0"), ]$large_r <- "None"
data_infl[which(data_infl$large_r == "1"), ]$large_r <- "Scarce"
data_infl[which(data_infl$large_r == "2"), ]$large_r <- "Rel. scarce"
data_infl[which(data_infl$large_r == "3"), ]$large_r <- "Abundant"
data_infl[which(data_infl$large_r == "4"), ]$large_r <- "Very abundant"

data_infl <- data_infl[order(-leverage), ]
names(data_infl) <- c("Source", "Area", "Classification",
                      "Minimum walking time from Bau (hr)", "Quality",
                      "Extent (m<sup>2</sup>)", "Large rock abundances",
                      "Leverage")

knitr::kable(data_infl, caption = "SOM Table S7. Characteristics (unscaled) of
             influential cases with leverage values above selected threshold,
             ordered by their leverage values")
```

### SOM S4 - Logistic model results:
A null model comparison revealed that our logistic model of the influence of the four tested variables on the likelihood of sources being classified as potentially utilized or not is highly significant ($\chi^2$=`r round(res_mod_sig$Deviance[2], 2)`,
df=`r res_mod_sig$Df[2]`, p `r report_pvals(res_mod_sig$P[2])`). As illustrated in
SOM Table S8 and as discussed below, the quality of the raw materials, the size of the area over which these are found, and the abundance of large nodules (more flexible in their potential for reduction), all have a positive and significant impact on the likelihood of sources being classified as potentially utilized.
However, the strongest
and most significant influence is that of the time it takes to reach a source from the Bau, and its effect is negative - the higher the access costs of a source, the less likely it is to be classified as potentially exploited. These results are consistent with previous findings (e.g., Browne and Wilson, 2011) as well as theoretical expectations.


```{r som-moddiag-infl-coefs}
# Test model with high leverage cases removed:
data_sel <- subset(glm_sel, hatvalues(mod) <= lev_thresh)
mod_sel <- glm(pot.used.fac ~ qual.log.z + ext.sqrt.z + time.sqrt.z + large_r.z,
               data = data_sel, family = "binomial")

mod_sum <- data.frame("Model" = paste("All cases (n=", length(mod$residuals),
                                      ")", sep = ""),
                      "Coefficient" = names(mod$coefficients),
                      "Estimate" = coef(summary(mod))[, "Estimate"],
                      "Std Error" = coef(summary(mod))[, "Std. Error"],
                      "Z value" = coef(summary(mod))[, "z value"],
                      "P" = coef(summary(mod))[, "Pr(>|z|)"])
mod_sum <- rbind(mod_sum,
                 data.frame("Model" = paste("High leverage cases removed (n=",
                                            length(mod_sel$residuals), ")",
                                            sep = ""),
                          "Coefficient" = names(mod_sel$coefficients),
                          "Estimate" = coef(summary(mod_sel))[, "Estimate"],
                          "Std Error" = coef(summary(mod_sel))[, "Std. Error"],
                          "Z value" = coef(summary(mod_sel))[, "z value"],
                          "P" = coef(summary(mod_sel))[, "Pr(>|z|)"]))
mod_sum[3:6] <- round(mod_sum[3:6], 3)
rownames(mod_sum) <- 1:nrow(mod_sum)
mod_sum <- knitr::kable(mod_sum, "html", caption = "SOM Table S8. Baseline scaled (z-transformed) logistic model coefficients with and without influential cases removed.")
kableExtra::collapse_rows(mod_sum, 1)
```


#### Minimum walking times from Bau {#SOM-walktimes}

``` {r}
time_probs <- predict_logistic_fits(mod, "time.sqrt.z",
                                   var_vals = glm_sel$time.sqrt.z, type = "raw")
```
<!-- NOTE: I'm using the means of the transformed variables here (i.e. as they
went into the model), NOT as they appear on their raw scale, since that's not the
value we're controlling for -->
SOM Figure S5 shows the influence of the cost of reaching a
source from the Bau on its likelihood of classification as potentially utilized,
when controlling for the effects of the other variables considered.
Sources of an average size (ca. `r round(mean(glm_sel$ext.sqrt)^2, 2)`m<sup>2</sup>), yielding
raw materials of average
quality for the region (ca. `r round(exp(mean(glm_sel$qual.log)), 2)`), and
characterized by an abundance of large rocks that is also typical for the study
area (i.e., absent or scarce), have a probability of being classified as
potentially utilized of about `r round(max(time_probs$fit), 2)*100`%
(`r round(time_probs[which.max(time_probs$fit), "lwr"], 2)*100`-`r round(time_probs[which.max(time_probs$fit), "upr"], 2)*100` at 95% confidence level)
<!-- Not using zero for time here because that would be extrapolating from the data -->
if in the vicinity of the Bau (i.e. where the closest source classified as
potentially utilized is located, or `r round(min(glm_sel$time_fromBau)/60, 1)`
minutes from the Bau). This probability falls below 50% if located at
more than about `r round(as.numeric(glm_sel[which.min(abs(time_probs$fit - .5)), c("time_fromBau")]/60))` minutes from the site. Note that these probabilities do not account for the spatial relationship between sources.


```{r som-modres-time, fig.cap="SOM Figure S5. Modeled probabilities of source classification (0 = unused; 1 = potentially utilized) based on the minimum walking times required to reach them from the from Bau, when all other source characteristics are kept at the mean values observed throughout the region. Actual observations are represented by black dots. Note that this variable has a major effect on classification probabilities, with estimates ranging from more than 0.75 with average sources located close to the site to almost zero with average sources located at ca. 13 hours from the Bau"}

## Walking time effect, all other predictors at their mean value:
res <- plot_binpred_inf(mod, data.frame(glm_sel), "time.sqrt.z",
                        "pot.used.fac", bin_freq = FALSE)
# Sort out labels:
res_brk_lbl <- 1:round(max(glm_sel$time_fromBau) / 3600) # Hours from the Bau.
# Transform to raw scale of the scaled variable
res_brk_trans <- sqrt(res_brk_lbl * 3600)
res_brk <- (res_brk_trans - mean(glm_sel$time.sqrt)) / sd(glm_sel$time.sqrt)
# Plot
p_time <- res$g + scale_x_continuous("Minimum walking time from Bau (hours)",
                                     breaks = res_brk, labels = res_brk_lbl) +
  labs(y = "Probability of classification as potentially used")
p_time
ggsave(filename = "../plots/figure_s5.jpg", plot = p_time, width = 20,
       height = 15, units = "cm", dpi = 600)
```

#### The influence of raw material quality {#SOM-rmqual}
``` {r}
qual_probs <- predict_logistic_fits(mod, "qual.log.z",
                                   var_vals = glm_sel$qual.log.z, type = "raw")
```
The influence of raw material quality on the likelihood of sources being classified
as potentially utilized when other variables are controlled for is shown in SOM
Figure S6. As indicated in subplot A, these likelihoods
are above 50% for sources with otherwise average characteristics for the region
(i.e. scarce large nodules and a surface area of roughly
`r round(mean(glm_sel$ext.sqrt)^2)`m<sup>2</sup>) and located in the vicinity
of the Bau (i.e. ca. `r round(min(glm_sel$time_fromBau)/60)` minutes) in all but
the lowest raw material quality cases. On the other hand, similar sources that
are only reachable in ca. `r round(mean(glm_sel$time.sqrt)^2/3600, 1)` hours or more from the Bau (the
adjusted mean for the region) would have to yield materials of the highest
quality (ca. `r round(exp(min(qual_probs[which(qual_probs$fit >= .5),]$qual.log.z) * sd(glm_sel$qual.log) + mean(glm_sel$qual.log)), 1)`) 
to have a greater than 50% probability of being classified as potentially utilized
(subplot B).

```{r som-modres-qual, fig.cap="SOM Figure S6. Modeled probabilities of source classification (0 = unused; 1 = potentially utilized) based on raw material quality, with minimum walking times (from the Bau) kept constant either at the observed minimum (A) or at the average for all sources in the region (B), and all other variables kept constant at their mean. Actual observations are represented by black dots whose size is proportional to the number of unused or potentially used sources with specific raw material quality values. Note that the presence of high quality raw materials plays an important role, but more so with sources that are not in the immediate vicinity of the site (B)."}
res <- plot_binpred_inf(mod, data.frame(glm_sel), "qual.log.z", "pot.used.fac",
                       vopt = "time.sqrt.z",
                       vopt_val = min(glm_sel$time.sqrt.z))
is.wholenumber <-
    function(x, tol = .Machine$double.eps^0.5)  abs(x - round(x)) < tol # From is.integer example
q_labs <- exp(res$xscale$labels)
q_labs_ni <- which(!is.wholenumber(q_labs))
q_labs[q_labs_ni[-1]] <- ""

p_qual_min <- res$g + scale_x_continuous("", breaks = res$xscale$breaks,
                                        labels = q_labs) +
  guides(size = FALSE) +
  labs(y = "Probability of classification as potentially used")



# Quality:
res <- plot_binpred_inf(mod, data.frame(glm_sel), "qual.log.z", "pot.used.fac")
# Note: Reusing var from previous plot, but the labels/breaks shouldn't change.
p_qual_avg <- res$g + scale_x_continuous("", breaks = res$xscale$breaks,
                                        labels = q_labs) +
  theme(axis.title.y = element_text(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "")

p_qual <- cowplot::plot_grid(p_qual_min, p_qual_avg, labels = "AUTO", nrow = 1)
p_qual_plt <- cowplot::ggdraw(cowplot::add_sub(p_qual, "Raw material quality",
                                 vpadding = grid::unit(0, "lines"),
                                 y = 6, x = 0.5, vjust = 4.5))
p_qual_plt

```

#### The influence of source extent {#SOM-extent}

For otherwise average sources located in the vicinity of the Bau, the size of the
area over which raw materials may be found somewhat shapes but certainly does
not fundamentally determine their probabilities of classification as potentially
utilized, which are in all cases above ca. 70% (SOM Fig. S7). This variable has
a somewhat clearer effect when access costs become a greater concern, but in all
cases the probability of classification as potentially utilized for otherwise
average sources (scarce large rocks, relatively low quality raw materials, ca. 3 hours away
from the Bau) is well below 50%. Assuming classification probabilities are adequate
proxies for selection probabilities, these data suggest that, when other criteria
were met, larger sources were preferred, possibly because
of a combination of factors such as being easier to find, easier to intercept
on the way back to the site, and easier to casually procure materials from (having to work less to find suitable stone). 

```{r som-modres-ext, fig.cap="SOM Figure S7. Modeled probabilities of source classification (0 = unused; 1 = potentially utilized) based on source extents, with minimum walking times (from the Bau) kept constant either at the observed minimum (A) or at the average for all sources in the region (B), and all other variables kept constant at their mean. Actual observations are represented by black dots whose size is proportional to the number of unused or potentially used sources with specific extent values. Note that source extent plays a role, with larger sources being more likely to be classified as potentially utilized, but the influence of this variable is relatively minor."}
res <- plot_binpred_inf(mod, data.frame(glm_sel), "ext.sqrt.z", "pot.used.fac",
                        vopt = "time.sqrt.z",
                        vopt_val = min(glm_sel$time.sqrt.z))
p_ext_min <- res$g + scale_x_continuous("", breaks = res$xscale$breaks,
                                        labels = round(res$xscale$labels ^ 2)) +
  guides(size = FALSE) +
  labs(y = "Probability of classification as potentially used")

res <- plot_binpred_inf(mod, data.frame(glm_sel), "ext.sqrt.z", "pot.used.fac")
p_ext_avg <- res$g + scale_x_continuous("", breaks = res$xscale$breaks,
                                        labels = round(res$xscale$labels ^ 2)) +
  theme(axis.title.y = element_text(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "")

p_ext <- cowplot::plot_grid(p_ext_min, p_ext_avg, labels = "AUTO", nrow = 1)
cowplot::ggdraw(cowplot::add_sub(p_ext, "Source extents (square meters)",
                                 vpadding = grid::unit(0, "lines"),
                                 y = 6, x = 0.5, vjust = 4.5))

```

#### The influence of large rock abundances {#SOM-larger}
As shown in SOM Figure S8, and as with source extent, the overall effect of this variable with otherwise
average sources located in the vicinity of the Bau is minor (above 50% in all
cases, and there is overlap in the 95% confidence intervals of the maximum
and minimum probabilities). For sources where access costs are greater the
influence of this variable becomes more critical. All else being equal, a
source which is average for the region in all respects (relatively low quality
materials found over a reasonably large area at ca. 3 hours from the Bau), has
a very low probability of being classified as potentially
utilized if it yields no or only few large rocks. However, for such a source the
effect of having an abundance of large rocks is far more uncertain, as shown by
95% confidence intervals which, for the maximum predicted probability, span almost
half of the possible range. If probabilities of classification can be
taken as reliable proxies for source selection, these data suggest,
overall, that when access costs were a concern sources with few large rocks were avoided,
while for sources where the abundance of large rocks was not an issue other
variables played a more important role in determining their likelihood of selection.


```{r som-modres-larger, fig.cap="SOM Figure S8. Modeled probabilities of source classification (0 = unused; 1 = potentially utilized) based on the abundance of large usable rocks, with minimum walking times (from the Bau) kept constant either at the observed minimum (A) or at the average for all sources in the region (B), and all other variables kept constant at their mean. Abundance values are given on a scale from 0 (absent) to 4 (very abundant; see also variable L in Table 1 of the main text). Actual observations are represented by black dots whose size is proportional to the number of unused or potentially used sources with specific abundance values. Note that the abundance of large rocks substantially and positively influences the likelihood estimates with sources that are not located in the immediate vicinity of the site (B)."}
res <- plot_binpred_inf(mod, data.frame(glm_sel), "large_r.z", "pot.used.fac",
                       vopt = "time.sqrt.z",
                       vopt_val = min(glm_sel$time.sqrt.z))
p_lr_min <- res$g + scale_x_continuous("", breaks = res$xscale$breaks,
                                       labels = res$xscale$labels) +
  guides(size = FALSE) +
  labs(y = "Probability of classification as potentially used")

res <- plot_binpred_inf(mod, data.frame(glm_sel), "large_r.z", "pot.used.fac")
p_lr_avg <- res$g + scale_x_continuous("", breaks = res$xscale$breaks,
                                       labels = res$xscale$labels) +
  theme(axis.title.y = element_text(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "")

p_lr <- cowplot::plot_grid(p_lr_min, p_lr_avg, labels = "AUTO", nrow = 1)
cowplot::ggdraw(cowplot::add_sub(p_lr, "Abundance of large rocks",
                                 vpadding = grid::unit(0, "lines"), y = 6,
                                 x = 0.5, vjust = 4.5))
```
